---
title: Merging the embryo chimera data set
author:
- name: Jonathan A. Griffiths
  affiliation: &CRUK Cancer Research UK Cambridge Institute, Li Ka Shing Centre, Robinson Way, Cambridge CB2 0RE, United Kingdom
- name: Aaron T. L. Lun
  affiliation: *CRUK
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{03. Embryo merging}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: ref.bib
---

<!--
AL: asterisks below remove weird highlighting on my editor.
****
-->

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
library(BiocStyle)
```

# Introduction

Once a processed `SingleCellExperiment` is available, the next step is to merge different samples to place them on the same coordinate system.
This allows us to perform downstream procedures like dimensionality reduction and clustering without needing to model sample-to-sample variation in expression.
We will merge using the mutual nearest neighbors (MNN) method [@haghverdi2017batch] as described `r Biocpkg("simpleSingleCell", "batch.html", "elsewhere")`.

```{r}
library(SingleCellExperiment)
sce <- readRDS("embryo_processed.rds")
```

# Feature selection

# Modelling the technical variability

We model the technical noise in each sample using the `multiBlockVar()` function as described `r Biocpkg("simpleSingleCell", "var.html#fitting-batch-specific-trends", "here")`.
As this data set does not contain spike-ins, we set `make.tech.trend=TRUE` to estimate the technical component based on Poisson noise -  
see `r Biocpkg("simpleSingleCell", "tenx.html#modelling-the-mean-variance-trend", "here")` for details.

```{r}
library(scran)
dec.out <- multiBlockVar(sce, block=sce$sample, 
    make.tech.trend=TRUE, weighted=FALSE)
dec.out[,-7] # don't show per-block results.
```

We also turn off weighting to ensure that each sample contributes equally to the results, regardless of the number of cells.
This is desirable here because different conditions may have different sets of highly variable genes (HVGs).
If one condition contains more cells, weighting by the number of cells would bias the combined statistics in favour of that condition's HVGs.
This differs from more typical applications of `multiBlockVar()` where all samples are replicates.
In such cases, the underlying set of HVGs should be similar and thus weighting could be used to improve precision without biasing the results.

The trend lies close to the lower bound of the per-gene variances for each sample (Figure \@ref(fig:trendplots)).
This indicates that the assumption of Poisson noise is reasonable.

```{r trendplots, fig.wide=TRUE, fig.cap="Mean-variance relationship in each sample, using statistics computed from log-normalized expression values. Each point represents a gene and the blue line represents the fitted Poisson trend in each sample."}
par(mfrow=c(2,2))
per.block.stats <- dec.out$per.block
for (i in colnames(per.block.stats)) {
    cur.stats <- per.block.stats[,i]
    plot(cur.stats$mean, cur.stats$total, xlab="Mean log-expression", 
        ylab="Variance", main=sprintf("Sample %s", i), pch=16)
    curve(metadata(cur.stats)$trend(x), add=TRUE, col="dodgerblue", lwd=2)
}
```

**Comments from Aaron:**

- Turning off weighting in `multiBlockVar()` (and indirectly, `combineVar()`) assumes that there are no samples with very few cells.
Such samples will have imprecise variance estimates that, without weighting, would add extra uncertainty to the combined statistics.

## Identifying genes of interest

We define the features of interest as those with net biological components greater than zero.
This enriches for genes that contain some biological variation, reducing the effect of uninteresting Poisson noise on downstream analyses.

```{r}
to.use <- dec.out$bio > 0
summary(to.use)
```

The injected KO cells were derived from a single male cell line while the host embryos were a mix of male or female mice.
Thus, there is a (largely uninteresting) sex effect between the WT and KO samples.
To mitigate this, we remove _Xist_ and genes on the Y chromosome.

```{r}
library(TxDb.Mmusculus.UCSC.mm10.ensGene)
loc <- mapIds(TxDb.Mmusculus.UCSC.mm10.ensGene, key=rowData(sce)$ENSEMBL, 
    keytype="GENEID", column="CDSCHROM")
is.y <- loc=="chrY" & !is.na(loc)
to.use <- to.use & !is.y 
to.use[which(rowData(sce)$SYMBOL=="Xist")] <- FALSE
sum(to.use)
```

We also remove the Tomato marker used to sort for KO cells.

```{r}
to.use[rowData(sce)$SYMBOL=="tomato-td"] <- FALSE
sum(to.use)
```

# Dimensionality reduction with PCA

We use principal components analysis (PCA) to perform dimensionality reduction prior to merging.
This reduces computational work and removes some high-dimensional noise, as `r Biocpkg("simpleSingleCell", "reads.html#denoising-expression-values-using-pca", "previously discussed")`.
We perform PCA across all samples by using `multiBatchPCA()` on our selected subset of genes in `to.use`.
This ensures each sample contributes equally to the definition of the coordinate space, as described `r Biocpkg("simpleSingleCell", "batch.html#hierarchical-merging", "here")`.

```{r}
library(batchelor)
library(BiocSingular)

set.seed(101) # for irlba.
pc.out <- batchelor::multiBatchPCA(sce, batch=sce$sample, 
    subset.row=to.use, get.variance=TRUE,
    BSPARAM=IrlbaParam(deferred=TRUE))
pc.out # one output matrix per level of 'sample'.
```

By default, `multiBatchPCA()` will return 50 PCs for all samples.
We use `denoisePCANumber()` to choose the number of PCs to retain based on our previous estimates of technical noise in this data set.
This discards later PCs until the variance lost is equal to the total technical component.

```{r}
to.retain <- denoisePCANumber(
    metadata(pc.out)$var.explained, # variance explained per PC.
    sum(dec.out$tech[to.use]), # technical noise in subset of genes.
    metadata(pc.out)$var.total # total variance in the data
)
to.retain
```

We then subset the matrices to only retain the first `to.retain` PCs in each sample.

```{r}
for (i in seq_along(pc.out)) {
    pc.out[[i]] <- pc.out[[i]][,seq_len(to.retain),drop=FALSE]
}
```

**Comments from Aaron:**

- The `BSPARAM=` argument instructs `multiBatchPCA()` to use methods from `r CRANpkg("irlba")` to speed up the PCA.
This is done by using an approximate algorithm, with deferred centered and scaling to preserve sparsity.
For large data sets, it is possible to achieve further speed gains with parallelization via the `BPPARAM=` argument; 
or by switching to randomized SVD in `r CRANpkg("rsvd")` via `BiocSingular::RandomParam()`.

# Batch correction with MNN

We use the `fastMNN()` function in a hierarchical manner as described `r Biocpkg("simpleSingleCell", "batch.html#hierarchical-merging", "elsewhere")`.
This involves merging the most similar samples before merging those that are more different, to weaken the assumption of shared populations required for correct MNN detection.
In this case, we first merge samples from the same genotype to remove the batch effect.
Note the use of `pc.input=TRUE` to specify that the input values are already in low-dimensional PC space.

```{r}
ko.out <- batchelor::fastMNN(pc.out[["1"]], pc.out[["2"]], 
    pc.input=TRUE, compute.variances=TRUE)
metadata(ko.out)$lost.var
wt.out <- batchelor::fastMNN(pc.out[["3"]], pc.out[["4"]], 
    pc.input=TRUE, compute.variances=TRUE)
metadata(wt.out)$lost.var
```

The `lost.var` represents the proportion of variance in each batch that is removed by the batch correction^[Specifically, the orthogonalization step, as described `r Biocpkg("simpleSingleCell", "batch.html#with-diagnostics", "here")`.].
If the assumptions underlying the MNN approach hold, this should be low and represent removal of noise along the batch vector.
A large proportion of lost variance (>10%) may be cause for concern as it suggests that biological structure within each batch is being discarded.

Our next step is to merge samples across genotypes to remove technical and uninteresting biological differences.
This includes sex and changes in expression induced by injection or cell sorting.

```{r}
overall <- batchelor::fastMNN(ko.out$corrected, wt.out$corrected,
    pc.input=TRUE, compute.variances=TRUE)
metadata(overall)$lost.var
```

We store the result as in the `reducedDims` slot of our `SingleCellExperiment` object, for use in downstream functions.

```{r}
reducedDim(sce, "corrected") <- overall$corrected
```

Note that the merge to create `overall` will also eliminate changes in expression caused by loss of _Tal1_.
This is necessary for the purposes of mapping all cells onto a common coordinate system.
Without such correction, cells of the same type would be separated by the genotype difference across samples, precluding common annotation in downstream analyses.
Nonetheless, differential expression upon KO is of substantial biological interest and will be recovered in our downstream differential testing.

# Visual inspection of merge results

We use $t$-stochastic neighbor embedding (t-SNE) plots [@van2008visualizing] to perform further dimensionality reduction for visualization.
In the uncorrected data, cells clearly separate by genotype with no visible batch effects between replicates (Figure \@ref(fig:beforeplot)).

```{r beforeplot, fig.cap="t-SNE plot of the embryo data before MNN correction. Each point represents a cell and is coloured by the sample of origin (KO: 1 and 2, WT: 3 and 4)."}
library(scater)
old <- sce
reducedDim(sce, "PCA") <- do.call(rbind, pc.out)
plotTSNE(sce, rerun=TRUE, run_args=list(use_dimred="PCA"), colour_by="sample")
```

After correction, cells from all samples are merged together in the majority of subpopulations (Figure \@ref(fig:afterplot)).
This is consistent with the removal of the inter-genotype differences and simplifies annotation and interpretation in downstream analyses.

```{r afterplot, fig.cap="t-SNE plot of the embryo data after MNN correction. Each point represents a cell and is coloured by the sample of origin."}
sce <- runTSNE(sce, use_dimred="corrected", colour_by="sample")
plotTSNE(sce, colour_by="sample")
```

# Concluding remarks

We save the `SingleCellExperiment` object with the merged coordinates to file.
This will be used for downstream analyses such as clustering and differential testing.

```{r}
saveRDS(sce, "embryo_merged.Rmd")
```

All software packages used in this workflow are publicly available from the Comprehensive R Archive Network (https://cran.r-project.org) or the Bioconductor project (http://bioconductor.org). 
The specific version numbers of the packages used are shown below, along with the version of the R installation.

```{r}
sessionInfo()
```

# References
